DCMIP2016: A Review of Non-hydrostatic Dynamical Core Design and Intercomparison of Participating Models

### 2017-05-01 DCMIP2016-Part1-Models.tex
more cosmetics regarding FVM parts of text

\begin{table}[p]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or non-hydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only three numerical methods are represented among participating models, namely finite-difference (FD), finite-volume (FV), and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, $\dot{\zeta}$, $T_v$, $p$, $q$ & Yin-Yang (\S \ref{sec:grid_yinyang}) & FD & C-grid  \\
ACME--A & H/NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
ICON & NH (D) & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
OLAM & NH (D) & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho \theta_{il}$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

The Finite-Volume Module (FVM) of the Integrated Forecasting System (IFS) is currently under development at 
ECMWF \citep{smolarkiewiczetalJCP2016,kuehnleinJCP2017,smolarkiewiczetalJCP2017}. FVM solves the 
compressible Euler equations in a geospherical framework \citep{szmelter2010,smolarkiewiczetalJCP2016}. 
A centered two-time-level semi-implicit 
integration scheme is employed with 3D implicit treatment of acoustic, buoyant, and rotational modes 
\citep{smolarkiewiczJCP2014}. The associated 3D Helmholtz problem is solved iteratively using a bespoke 
preconditioned Generalised Conjugate Residual approach. The integration procedure uses the 
non-oscillatory MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) 
advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
The horizontal spatial discretization is fully unstructured finite-volume using the median-dual approach. This is 
combined with a structured-grid finite-difference discretization in the vertical direction. In both the horizontal and 
the vertical discretization all variables are co-located. 
The median-dual finite-volume mesh in the horizontal is developed about the nodes of the octahedral reduced 
Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral reduced Gaussian 
grid is also employed in the spectral-transform dynamical core of the presently operational IFS at ECMWF, which facilitates 
interoperability of the two formulations. However, FVM is not restricted to this grid and offers capabilities towards 
broad classes of meshes including adaptivity, e.g.~\cite{szmelter2010,kuehnleinJCP2012}.
%No explicit diffusion or filtering is applied in FVM for DCMIP, unless it is required for particular test cases.

@@

As with the classical reduced Gaussian grid of \cite{hortal1991use}, the octahedral 
reduced Gaussian grid \citep{malardel2016new,smolarkiewiczetalJCP2016} specifies the latitudes according to the 
roots of the Legendre polynomials. 
The two grids differ in the arrangement of the points along the latitudes, which follows a simple
rule for the octahedral grid: starting with 20 points on the first latitude around the poles, four points 
are added with every latitude towards the equator, whereby the spacing between points along the 
latitudes is uniform and there are no points at the equator. The octahedral reduced Gaussian grid 
is suitable for transformations involving spherical harmonics, and has been introduced for global 
numerical weather prediction with the spectral dynamical core of the IFS at ECMWF in 2016. 
Figure~\ref{fig:octahedralgrid} depicts the octahedral reduced Gaussian grid nodes together with
the edges of the primary mesh as applied in the context of the finite-volume discretisation of FVM (Section~\ref{sec:FVM}).

\begin{figure}
\centering
\begin{subfigure}{.4\textwidth}
  \centering
  \includegraphics[width=6cm]{contributions/FVM/octahedral_gauss_nodes}
 \end{subfigure}%
\begin{subfigure}{.4\textwidth}
  \centering
  \includegraphics[width=5.4cm]{contributions/FVM/octahedral_RGG_white}
 \end{subfigure}
\caption{Locations of the octahedral reduced Gaussian grid nodes (left), and the edges of the primary mesh 
connecting the nodes as applied with the finite-volume discretisation in FVM (right). A coarse octahedral grid with only
24 latitudes between pole and equator 'O24' is used for illustration. The dual mesh resolution of the octahedral
reduced Gaussian grid is about a factor 2 finer at the poles than the equator; see \cite{smolarkiewiczetalJCP2016}.}
\label{fig:octahedralgrid}
\end{figure}

@@

The FVM formulation is based on conservation laws for dry mass \eqref{fvm:mass}, momentum \eqref{fvm:momentum}, 
and dry entropy \eqref{fvm:thermodynamic} in Eulerian flux-form,
which are similar to \eqref{eq:ContinuityEquation} for $\rho_d$, \eqref{eq:PrognosticVelocityFluxForm}, and 
\eqref{eq:PotentialTemperatureFluxForm} for $\theta$, respectively. Moreover, underlying the conservation laws 
in FVM is a perturbational form with respect to a balanced ambient state and a generalized curvilinear 
coordinate formulation in a geospherical framework. Following \cite{smolarkiewiczetalJCP2017}, the
governing equations can concisely be written as
\begin{subequations} \label{fvm:compressible}
\begin{align}
 & \pdiff{\mathcal{G}\rho_d}{t} + \overline{\nabla} \cdot \left(\mathbf{v} \mathcal{G} \rho_d\right) = 0~,
\label{fvm:mass}  \\
& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \overline{\nabla} \cdot \left(\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
\mathcal{G}\rho_d\left( - \theta_{\rho} \mathbf{\widetilde{G}} \overline{\nabla} \phi' 
- \mathbf{k} \frac{g_c}{\theta_{a}} \left(\theta'+\theta_a \left(\frac{q'_v}{\varepsilon} - q'_v - q_c - q_r \right) \right)
- 2\mathbf{\Omega} \times \left( \mathbf{u} - \frac{\theta_{\rho}}{\theta_{\rho\,a}} \mathbf{u}_{a} \right) 
+ \mathcal{M}' \right]~,
%+ \mathbf{M} \right)~,
\label{fvm:momentum} \\
%- \frac{L}{c_p \pi} & \Big( \frac{\Delta q_{vs}}{\Delta t} & + E_r  \Big)
& \pdiff{\mathcal{G}\rho_d \theta'}{t} + \overline{\nabla} \cdot \left(\mathbf{v}\mathcal{G}\rho_d\,\theta' \right) = 
- \mathcal{G}\rho_d\,\mathbf{\widetilde{G}}^{T}\mathbf{u} \cdot \overline{\nabla} \theta_{a} ~,
%- \frac{L}{c_{pd}\,\pi} \left( \frac{\Delta q_{vs}}{\Delta t}  + E_r  \right) %\left(C_d + E_p\right) 
%+  \mathcal{H} \right)~,
\label{fvm:thermodynamic} \\
& \phi' = c_{pd} \left[\left( \frac{R_{d}}{p_{0}} \rho_d\,\theta\,(1+q_v/\varepsilon) \right)^{R_{d}/c_{vd}} - \pi_{a} \right]~.
\label{fvm:gaslaw}
\end{align}
\end{subequations}
Dependent variables in \eqref{fvm:compressible} are dry density $\rho_d$, 3D
physical velocity vector $\mathbf{u}$, potential temperature perturbation $\theta'$, and a modified
Exner pressure perturbation $\phi'$, 
with the thermodynamic variables related by the gas law 
\eqref{fvm:gaslaw}. Primes indicate perturbations with respect to the prescribed ambient state denoted by
subscript `a', see \cite{prusa2008eulag,smolarkiewiczJCP2014} for discussion. As far as geometric aspects 
are concerned, the nabla operator $\overline{\nabla}$ represents
the 3D vector of partial derivatives with respect to the curvilinear coordinates, along with the Jacobian $\mathcal{G}$, 
a matrix of metric coefficients $\mathbf{\widetilde{G}}$, its transpose $\widetilde{\mathbf{G}}^{T}$, and 
the contravariant velocity $\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$ where a contribution from
optional time-dependency of the curvilinear coordinates is neglected for simplicity \citep{kuehnleinJCP2017}.
The symbol $\mathcal{M}' = \mathcal{M}'(\mathbf{u}, \mathbf{u}_a, \theta_{\rho}/\theta_{\rho\,a})$ in \eqref{fvm:momentum} 
subsumes the metric forces in the spherical domain, see \cite{smolarkiewiczetalJCP2017}.

@@

A characteristic feature of the FVM (Section~\ref{sec:FVM}) time-stepping scheme is the 3D implicit treatment 
of the fast buoyant and acoustic modes, and the slow rotational modes. Therefore, the model time step is 
identical for all processes and typically selected with regard to the stability of the advective transport scheme---i.e.~%
the time step is continuously adapted according to a given maximum advective Courant number 
permitted by the MPDATA scheme.
A comprehensive discussion of the integration scheme can be found in \cite{smolarkiewiczJCP2014,smolarkiewiczetalJCP2016,kuehnleinJCP2017} for dry dynamics, 
and in \cite{kurowskiJAS2014,smolarkiewiczetalJCP2017} for extension to moist-precipitating dynamics.  
Here, we provide a short outline of the solution procedure for the compressible Euler 
equations \eqref{fvm:compressible}. It employs the two-time-level second-order-accurate template algorithm given as
\begin{equation}
\psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\psi}^{n},\mathbf{V}^{n+1/2}, (\mathcal{G}\rho_d)^{n}, (\mathcal{G}\rho_d)^{n+1})
+0.5\,\Delta t\,R^{\psi}|_{\mbf{i}}^{\scs n+1}~,
\hspace{1em} \widetilde{\psi}^{n}\,\equiv\,\psi^n+0.5\,\Delta t\,R^{\psi}|^{\scs n}~.
\label{fvm:solscalar}
\end{equation}
where $\psi$ represents the solution variable, $R^{\psi}$ is the respective rhs,
$\mathcal{A}$ symbolises the advective transport operator given by the non-oscillatory finite-volume 
MPDATA  scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}, and the spatial mesh vector 
index $\mathbf{i} \equiv (k,i)$ denotes the position on the hybrid horizontally-unstructured 
vertically-structured computational mesh.

The solution procedure of the system \eqref{fvm:compressible} can then basically be divided into three steps. First, the homogenous
mass continuity equation \eqref{fvm:mass} is integrated with $\psi \equiv \rho_d$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G} $,
and $R^{\rho_d} \equiv 0$ in \eqref{fvm:solscalar}. Second, given already updated moisture variables 
\citep{smolarkiewiczetalJCP2017}, the thermodynamic \eqref{fvm:thermodynamic} and momentum \eqref{fvm:momentum} equations enter
\eqref{fvm:solscalar} with $\psi = u, v, w, \theta'$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G}\rho_d$, and the rhs $R^{\psi}$ 
which is generally depending on all these prognostic variables. The high degree of implicitness in the representation of the rhs forcings
is achieved by inverting the overall discrete system \eqref{fvm:solscalar} to obtain closed-form expressions for the
velocity updates---this procedure is facilitated by the co-located arrangement of all variables on the computational mesh.
Retained on the rhs of the derived closed-form velocity expressions is the pressure gradient term.
The subsequent third step in the solution procedure is to formulate an implicit boundary value problem for the pressure variable $\phi'$ using 
an advective form of the equation of state \eqref{fvm:gaslaw}. An $\mathcal{O}(\Delta t^2)$ integration of this 
equation with a Euler backward scheme, in the spirit of \eqref{fvm:solscalar}, leads a Helmholtz equation
\citep{smolarkiewiczJCP2014}.
The associated 3D elliptic boundary value problem is solved iteratively using a preconditioned Generalised Conjugate 
Residual approach; see \cite{smolarkiewiczAG2011} for a recent overview and comprehensive list of references.
Nonlinear terms in $R^{\psi}|^{\scs n+1}$ and the 
solution-dependent coefficients of the Helmholtz equation are lagged behind and executed in an outer iteration.

>>>>

\begin{table}[p]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or non-hydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only three numerical methods are represented among participating models, namely finite-difference (FD), finite-volume (FV), and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q_v$ & Octahedral (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, $\dot{\zeta}$, $T_v$, $p$, $q$ & Yin-Yang (\S \ref{sec:grid_yinyang}) & FD & C-grid  \\
ACME--A & H/NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
ICON & NH (D) & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
OLAM & NH (D) & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho \theta_{il}$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

The Finite-Volume Module (FVM) of the Integrated Forecasting System (IFS) is currently under development at 
ECMWF \citep{smolarkiewiczetalJCP2016,kuehnleinJCP2017,smolarkiewiczetalJCP2017}. FVM solves the 
compressible Euler equations in a geospherical framework \citep{szmelter2010,smolarkiewiczetalJCP2016}. 
A centered two-time-level semi-implicit 
integration scheme is employed with 3D implicit treatment of acoustic, buoyant, and rotational modes 
\citep{smolarkiewiczJCP2014}. The associated 3D Helmholtz problem is solved iteratively using a bespoke 
preconditioned Generalised Conjugate Residual approach. The integration procedure uses the 
non-oscillatory MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) 
advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
The horizontal spatial discretization is fully unstructured finite-volume using the median-dual approach. This is 
combined with a structured-grid finite-difference discretization in the vertical direction. In both the horizontal and 
the vertical discretization all variables are co-located. 
The median-dual finite-volume mesh in the horizontal is developed about the nodes of the octahedral reduced 
Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral reduced Gaussian 
grid is also employed in the spectral-transform dynamical core of the presently operational IFS at ECMWF, which facilitates 
interoperability of the two formulations. However, FVM is not restricted to this grid and offers capabilities towards 
broad classes of meshes including adaptivity \citep[e.g.][]{szmelter2010,kuehnleinJCP2012}.
%No explicit diffusion or filtering is applied in FVM for DCMIP, unless it is required for particular test cases.

@@

As with the classical reduced Gaussian grid of \cite{hortal1991use}, the octahedral 
reduced Gaussian grid \citep{malardel2016new,smolarkiewiczetalJCP2016} specifies the latitudes according to the 
roots of the Legendre polynomials. 
The two grids differ in the arrangement of the points along the latitudes, which follows a simple
rule for the octahedral grid: starting with 20 points on the first latitude around the poles, four points 
are added with every latitude towards the equator, whereby the spacing between points along the 
latitudes is uniform and there are no points at the equator. The octahedral reduced Gaussian grid 
is suitable for transformations involving spherical harmonics, and has been introduced for operational 
weather prediction with the spectral dynamical core of the IFS at ECMWF in 2016. 
Figure~\ref{fig:octahedralgrid} depicts the octahedral reduced Gaussian grid nodes together with
the edges of the primary mesh as applied in the context of the finite-volume discretisation of FVM (Section~\ref{sec:FVM}).

\begin{figure}
\centering
\begin{subfigure}{.4\textwidth}
  \centering
  \includegraphics[width=6cm]{contributions/FVM/octahedral_gauss_nodes}
 \end{subfigure}%
\begin{subfigure}{.4\textwidth}
  \centering
  \includegraphics[width=5.4cm]{contributions/FVM/octahedral_RGG_white}
 \end{subfigure}
\caption{Locations of the octahedral reduced Gaussian grid nodes (left), and the edges of the primary mesh 
connecting the nodes as applied with the finite-volume discretisation in FVM (right). A coarse octahedral grid with only
24 latitudes between pole and equator 'O24' is used for illustration. The dual mesh resolution of the octahedral
reduced Gaussian grid is about a factor 2 finer at the poles than the equator, see \cite{smolarkiewiczetalJCP2016}.}
\label{fig:octahedralgrid}
\end{figure}

@@

The FVM formulation is based on conservation laws for dry mass \eqref{fvm:mass}, momentum \eqref{fvm:momentum}, 
and dry entropy \eqref{fvm:thermodynamic} in Eulerian flux-form,
which are similar to \eqref{eq:ContinuityEquation} for $\rho_d$, \eqref{eq:PrognosticVelocityFluxForm}, and 
\eqref{eq:PotentialTemperatureFluxForm} for $\theta$, respectively. Moreover, underlying the conservation laws 
in FVM is a perturbational form with respect to a balanced ambient state and a generalized curvilinear 
coordinate formulation in a geospherical framework. Following \cite{smolarkiewiczetalJCP2017}, the
FVM governing equations can concisely be written as
\begin{subequations} \label{fvm:compressible}
\begin{align}
 & \pdiff{\mathcal{G}\rho_d}{t} + \overline{\nabla} \cdot \left(\mathbf{v} \mathcal{G} \rho_d\right) = 0~,
\label{fvm:mass}  \\
& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \overline{\nabla} \cdot \left(\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
\mathcal{G}\rho_d\left( - \theta_{\rho} \mathbf{\widetilde{G}} \overline{\nabla} \phi' 
%- \mathbf{k} \frac{g}{\theta_{a}} \left(\theta'+\theta_a \left(\frac{q'_v}{\varepsilon} - q'_v - q_c - q_r \right) \right)
%- \mathbf{k}\,g \left(\frac{\theta'}{\theta_a}+\frac{q'_v}{\varepsilon} - q'_v - q_c - q_r \right)
- \mathbf{k}\,g \left(\frac{\theta'}{\theta_a}+\varepsilon_b\,q'_v - q_c - q_r \right)
- 2\mathbf{\Omega} \times \left( \mathbf{u} - \frac{\theta_{\rho}}{\theta_{\rho\,a}} \mathbf{u}_{a} \right) 
+ \mathcal{M}' \right]~,
%+ \mathbf{M} \right)~,
\label{fvm:momentum} \\
%- \frac{L}{c_p \pi} & \Big( \frac{\Delta q_{vs}}{\Delta t} & + E_r  \Big)
& \pdiff{\mathcal{G}\rho_d \theta'}{t} + \overline{\nabla} \cdot \left(\mathbf{v}\mathcal{G}\rho_d\,\theta' \right) = 
- \mathcal{G}\rho_d\,\mathbf{\widetilde{G}}^{T}\mathbf{u} \cdot \overline{\nabla} \theta_{a} ~,
%- \frac{L}{c_{pd}\,\pi} \left( \frac{\Delta q_{vs}}{\Delta t}  + E_r  \right) %\left(C_d + E_p\right) 
%+  \mathcal{H} \right)~,
\label{fvm:thermodynamic} \\
& \phi' = c_{pd} \left[\left( \frac{R_{d}}{p_{0}} \rho_d\,\theta\,(1+q_v/\varepsilon) \right)^{R_{d}/c_{vd}} - \pi_{a} \right]~.
\label{fvm:gaslaw}
\end{align}
\end{subequations}
Dependent variables in \eqref{fvm:compressible} are dry density $\rho_d$, 3D
physical velocity vector $\mathbf{u}$, potential temperature perturbation $\theta'$, and a modified
Exner pressure perturbation $\phi'$, 
with the thermodynamic variables related by the gas law 
\eqref{fvm:gaslaw}. Primes indicate perturbations with respect to the prescribed ambient state denoted by
subscript `a', see \cite{prusa2008eulag} and \cite{smolarkiewiczJCP2014} for discussions.
The symbol $g$ in \eqref{fvm:momentum} denotes the gravitational acceleration 
and $\varepsilon_b = 1/\varepsilon-1$. As far as geometric aspects 
are concerned, the nabla operator $\overline{\nabla}$ represents
the 3D vector of partial derivatives with respect to the curvilinear coordinates, along with the Jacobian $\mathcal{G}$, 
a matrix of metric coefficients $\mathbf{\widetilde{G}}$, its transpose $\widetilde{\mathbf{G}}^{T}$, and 
the contravariant velocity $\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$ where a contribution from
optional time-dependency of the curvilinear coordinates is neglected for simplicity \citep{kuehnleinJCP2017}.
The symbol $\mathcal{M}' = \mathcal{M}'(\mathbf{u}, \mathbf{u}_a, \theta_{\rho}/\theta_{\rho\,a})$ in \eqref{fvm:momentum} 
subsumes the metric forces in the spherical domain \citep{smolarkiewiczetalJCP2017}.

@@

A characteristic feature of the FVM (Section~\ref{sec:FVM}) time-stepping scheme is the 3D implicit treatment 
of the fast buoyant and acoustic modes, and the slow rotational modes. Therefore, the model time step is 
identical for all processes and typically selected with regard to the stability of the advective transport scheme---i.e.~%
the time step is continuously adapted according to a given maximum advective Courant number 
permitted by the MPDATA scheme.
A comprehensive discussion of the integration scheme can be found in \cite{smolarkiewiczJCP2014,smolarkiewiczetalJCP2016}
and \cite{kuehnleinJCP2017} for dry dynamics, 
whereas in \cite{kurowskiJAS2014} and \cite{smolarkiewiczetalJCP2017} for extension to moist-precipitating dynamics.  
Here, we provide a short outline of the solution procedure for the compressible Euler 
equations \eqref{fvm:compressible}. It employs the two-time-level second-order-accurate template algorithm given as
\begin{equation}
\psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\psi}^{n},\mathbf{V}^{n+1/2}, (\mathcal{G}\rho_d)^{n}, (\mathcal{G}\rho_d)^{n+1})
+0.5\,\Delta t\,R^{\psi}|_{\mbf{i}}^{\scs n+1}~,
\hspace{1em} \widetilde{\psi}^{n}\,\equiv\,\psi^n+0.5\,\Delta t\,R^{\psi}|^{\scs n}~.
\label{fvm:solscalar}
\end{equation}
where $\psi$ represents the solution variable, $R^{\psi}$ is the respective rhs,
$\mathcal{A}$ symbolises the advective transport operator given by the non-oscillatory finite-volume 
MPDATA  scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}, and the spatial mesh vector 
index $\mathbf{i} \equiv (k,i)$ denotes the position on the hybrid horizontally-unstructured 
vertically-structured computational mesh.

The solution procedure of the system \eqref{fvm:compressible} can then basically be divided into three steps. First, the homogenous
mass continuity equation \eqref{fvm:mass} is integrated with $\psi \equiv \rho_d$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G} $,
and $R^{\rho_d} \equiv 0$ in \eqref{fvm:solscalar}. Second, given already updated moisture variables 
\citep{smolarkiewiczetalJCP2017}, the thermodynamic \eqref{fvm:thermodynamic} and momentum \eqref{fvm:momentum} equations enter
\eqref{fvm:solscalar} with $\psi = u, v, w, \theta'$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G}\rho_d$, and the rhs $R^{\psi}$ 
which is generally depending on all these prognostic variables. The high degree of implicitness in the representation of the rhs forcings
is achieved by inverting the overall discrete system \eqref{fvm:solscalar} to obtain closed-form expressions for the
velocity updates---this procedure is facilitated by the co-located arrangement of all variables on the computational mesh.
Retained on the rhs of the derived closed-form velocity expressions is the pressure gradient term.
The subsequent third step in the solution procedure is to formulate an implicit boundary value problem for the pressure variable $\phi'$ using 
an advective form of the equation of state \eqref{fvm:gaslaw}. An $\mathcal{O}(\Delta t^2)$ integration of this 
equation with a Euler backward scheme, in the spirit of \eqref{fvm:solscalar}, leads a Helmholtz equation
\citep{smolarkiewiczJCP2014}.
The associated 3D elliptic boundary value problem is solved iteratively using a preconditioned Generalised Conjugate 
Residual approach, see \cite{smolarkiewiczAG2011} for a recent overview and comprehensive list of references.
Nonlinear terms in $R^{\psi}|^{\scs n+1}$ and the 
solution-dependent coefficients of the Helmholtz equation are lagged behind and executed in an outer iteration.
